Research Notes

### 2018-07-30 writeups/notes.tex
NS solver
check in modulational instability writeup

>>>>

\section{Algebraic Manipulations}

\subsection{Factoring out the exponential dependence}

I've had to do this algebra too many times now, so let's just consider if we
factor out the $e^{z/2H}$ exponential scaling on the dynamical variables and the
$e^{-z/H}$ scaling on $\rho_0$. This leaves us with:
\begin{subequations}\label{se:nonlin_factored}
    \begin{align}
        \vec{\nabla} \cdot \vec{u}_1 + \frac{u_{1z}}{2H} &= 0,\\
        \pd{\rho_1}{t} - \frac{u_{1z} \rho_0}{H} &= -e^{z/2H}\s*{
            \p*{\vec{u}_1 \cdot \vec{\nabla} \rho_1} + \frac{u_{1z}\rho_1}{2H}
        },\\
        \pd{\vec{u}_1}{t} + \frac{\vec{\nabla}P}{\rho_0}
            + \frac{P}{2H\rho_0}\hat{z} + \frac{\rho_1\vec{g}}{\rho_0}
            &= -e^{z/2H}\s*{
            \p*{\vec{u} \cdot \vec{\nabla} \vec{u}_1}
                + \frac{u_{1z}\vec{u}_1}{2H}
        }.
    \end{align}
\end{subequations}
I'm too lazy to re-notate my variables, so all of them are constant amplitude
over $z$ now.

\chapter{Stability Analysis}

\section{Modulational Instability of NLSE}

We study the nonlinear Schr\"odinger Equation (NLSE)
\begin{equation}
    i\psi_t + \psi_{xx} + \nu \abs*{\psi}^2\psi = 0.
\end{equation}
Sutherland's paper demonstrates an interaction between the wave and the
wave-induced mean flow ($\propto \abs*{\psi}^2$) that reduces the IGW system to
the NLSE\@. Here we follow Whitham's textbook in rederiving the modulational
instability criterion.

Consider for some fixed value $k_0$, then the linear SE (LSE) admits a plane
wave solution $\psi = e^{i(k_0x - \omega(k_0)t)}$ for some dispersion relation
$\omega(k)$. If instead we permit $\psi$ to be distributed in frequency space
near $k_0$, then we may notate
\begin{equation}
    \psi = \psi_0 \varphi = e^{i(k_0x - \omega(k_0) t)} \int\limits
        \exp\s*{i\p*{k' x - k'\pd{\omega}{k}t -
            \frac{1}{2}k'^2 \ptd{\omega}{k}t}}\;\mathrm{d}k',
\end{equation}
where $k' = k - k_0$ denotes the deviation from the central $k_0$
wavenumber. Defining $\omega' = \omega - \omega_0$, we may observe
\begin{equation}
    \omega' = k' \pd{\omega}{k} + \frac{1}{2}k'^2 \ptd{\omega}{k}.
\end{equation}
Making correspondence $-i\omega' \to \partial_t, ik' \to \partial_x$, we see
that $\varphi$ obeys dispersion relation
\begin{equation}
    i\p*{\varphi_t + \pd{\omega}{k}\varphi_x}
        + \frac{1}{2}\ptd{\omega}{k}\varphi_{xx} = 0.
\end{equation}
Thus, the modulation $\varphi$ translates at the group velocity, and we find
that $\varphi$ obeys a modified LSE in a comoving frame.

If we reintroduce the nonlinearity, we observe that $\nu \abs*{\psi}^2\psi = \nu
\abs*{\varphi}^2 \psi_0\varphi$, and so the nonlinearity modifies the LSE as
\begin{align}
    i\p*{\varphi_t + \pd{\omega}{k}\varphi_x}
        + \frac{1}{2}\ptd{\omega}{k}\varphi_{xx}
        + \nu \abs*{\varphi^2}\varphi &= 0,\\
    \omega' = k' \pd{\omega}{k} + \frac{1}{2}k'^2 \ptd{\omega}{k}
        - \nu a^2.\label{eq:nlse_disp}
\end{align}

Armed with this dispersion relation, we can apply the method of characteristics
to the system of equations describing any wavetrain
\begin{subequations}\label{eq:wavetrain}
    \begin{align}
        \pd{k}{t} + \pd{\omega}{x} &= 0,\label{eq:wavetrain_1}\\
        \pd{A^2}{t} + \pd{}{x}\p*{c_gA^2} &= 0.\label{eq:wavetrain_2}
    \end{align}
\end{subequations}
We denote $A$ the wave amplitude and $c_g$ the group velocity. In a nonlinear
system, $\omega = \omega(A, k)$, and to leading order in $A$ we expect $\omega =
\omega_0 + \omega_2A^2$, i.e.\ the dispersion should not depend on the sign of
$A$. Thus we may rewrite \autoref{eq:wavetrain} as
\begin{subequations}\label{eq:wavetrain_mod}
    \begin{align}
        \pd{k}{t} + \p*{\pd{\omega_{0}}{k} + \pd{\omega_2}{k}A^2}\pd{k}{x}
            + \omega_2 \pd{A^2}{x} &= 0,\\
        \pd{A^2}{t} + \pd{A^2}{x}\pd{\omega_0}{k} + A^2\ptd{\omega_0}{k}\pd{k}{x}
            &= 0.
    \end{align}
\end{subequations}
We first note that, in $A \to 0$, the equations decouple and both equations
describe propagation at the group velocity (they are already in characteristic
form $Q_{,t} + cQ_{,x} = 0$). With some foresight, we anticipate that the
$\pd{A^2}{x}$ term in the first equation will provide a $\mathcal{O}(A)$
correction to the group velocity, while the $\pd{\omega_2}{k}A^2\pd{k}{x}$ term
provides an $\mathcal{O}(A^2)$ correction, so we drop it.

The method of characteristics then requires we seek linear combinations of the
two equations such that all equations are in characteristic form. The Whitham
prescription says that if the original system is of form $A_{ij}q_{j, t} +
B_{ij}q_{j, x} = 0$, then the characteristic curves $X(T)$ satisfy $A_{ij}X +
B_{ij}T = 0$. Here, this yields
\begin{equation}
    X(T) = \p*{\pd{\omega_0}{x} \pm A\sqrt{\omega_2 \ptd{\omega_0}{k}}}T.
\end{equation}

Comparing to the NLSE dispersion relation \autoref{eq:nlse_disp}, we identify
that $\sqrt{-q \ptd{\omega_0}{k}}A$ is the correction to the group velocity.
Thus, if $q\ptd{\omega_0}{k} > 0$, we obtain a \emph{complex} group velocity,
which is identified with \emph{modulational instability}. Naively, this might
imply a steepening $k = c_{g, 0}t - i\sqrt{q\omega_0''}t$ with timescale
$\Gamma \sim A\p*{q\omega_0''}^{-1/2}$.
